Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-19 paper-1/manuscript.tex
Rev sec Results:STD, F8 partial

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than other responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.
To show this in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the influence of both decay and unrelated activity in the intervening trial. In fact, even without intervening activity, inputs to $x_j$ would decay to a negligible $e^\frac{1 s}{\tau_x} = 0.0013$ within two trials.
Thus, STD could not have any stimulus-specific effects across dev target trials or msc trials.

However, given the substantial correlation from one trial to the next, the frequently presented stimulus B could have experienced stimulus-specific adaptation as a result of STD. Referring back to Figure \ref{fig3}b, we see that there was indeed a significant, if small, effect of presentation frequency even in the ablated model with STD only, yielding a positive SSA index on average.

To show that STD carry-over was responsible for the reduction of non-target dev pulses relative to msc, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) is affected by changes in STD. Unlike $V_{TA}$, which directly affects the neuron to which the value belongs, $x_j$ exerts its effects at the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost. $D$ is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at rest (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, ignoring added depression due to repeated spikes within a trial, and instead using the average depression at the start of trials $x_j^S$, we estimated the average peak EPSP evoked by a presynaptic neuron $j$ as $R_j^S k x_j^S$. Finally, with $x_j \in [0,1]$ and summing over presynaptic neurons, we could estimate the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate.

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} Estimated STD-mediated depression $D$ has small direct effects.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right). The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Statistics across networks and stimuli of the contrast (dev - msc) in $D$ ($D^B$ is shown first, since all other data in this panel relate directly to A), and in $V_{TA}$ at the start of A trials (cf. Figure \ref{fig7}d). In each network, averages were taken over neurons active (>0.2 spikes/trial) in either dev or msc trials (B trials for $\Delta D^B$, A trials for $\Delta D^A$ and $\Delta V_{TA}$). Whereas the third box ($\Delta V_{TA}$ total) indicates averages in the full model, the final box ($\Delta V_{TA}$ added) indicates the difference between the full and the ablated (TA only) model

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before. 
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the network in both conditions. While there were differences between the two conditions, they were relatively small and did not appear to follow a readily identifiable pattern. Conversely, B trials (bottom) were spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response size: The context was the same, but the B response tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site (see also Figure \ref{fig6}a), with depression, of course, similarly limited. As a result, the contrast (dev - msc) shows slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicative of increased synaptic depression in dev.

>>>>

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than other responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.

We had good reasons for this assumption: Firstly, notice that the ablated model with STD only produced only a positive SSA index (i.e., the target response in msc was greater than in std), but not a significantly positive DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B could lead to a build-up of depression strong enough to suppress its own response. Referring back to Figure \ref{fig3}b, we see that there was indeed a significant, if small, effect of presentation frequency even in the ablated model with STD only, yielding a positive SSA index on average.
% To show this in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the influence of both decay and unrelated activity in the intervening trial.

To show that STD being carried over between trials was responsible for the reduction of non-target dev pulses relative to msc, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $V_{TA}$, which directly affected the neuron to which the value belongs, $x_j$ exerted its effects on the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring the effect of $V_{TA}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to bring a neuron to fire.
A depression measure is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at rest (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $R_j^S k x_j^S$. For this estimate, we ignored added depression due to repeated spikes within a trial, instead defining $x_j^S$ as the average depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate. Lastly, since we contrasted this measure between conditions (see below), we aimed to avoid effects purely due to differences in presynaptic activity. Therefore, we calculated $R^S$ as the average response across relevant trials in all conditions, leaving the contrast entirely in the hands of $x^S$.

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} Estimated STD-mediated depression $D$ has small direct effects.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right). The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Statistics across networks and stimuli of the contrast (dev - msc) in $D$ in A trials (left) and B trials (right)

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) were spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response size: The context was the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site (see also Figure \ref{fig6}a), with depression by definition similarly limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicative of greatly increased synaptic depression in dev.
